---
title: "mcfly"
author: "Leandro Duarte, Gabriel Nakamura"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mcfly - an R package to estimate the influence of environmental niche selection on species     diversity in metacommunities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

#Introduction to **mcfly** package

Mcfly package was designed to estimate the influence of stabilizing selection acting along the macroevolutionary dynamics of a given lineage on niche-based species sorting accross sets of local communities, and its consequence for species diversity gradients.
We demonstrate in this vignette all steps needed to carry on analysis using the mcfly function. First of all, we need to install and read the mcfly package from github repository:

```{r setup, eval= FALSE, echo= TRUE}
library(devtools)
install_github("sokole/MCSim")
install_github("GabrielNakamura/mcfly")
library(mcfly)
```

##Data: Furnariidae family

We will run all the analysis using occurrence data of avian species from Furnariidae family. This data is part of the mcfly package. To read it, type:
```{r readData, eval=FALSE}
data("Furnariidae")
```
Furnariidae dataset comprises the occurence of 223 Furnariidae species distributed across the entire Neotropical region. The phylogenetic hypothesis used in this analysis was obtained from [BirdTree](https://birdtree.org/) and are presented in phylo_Furnariidae data, also presented in mcfly package.
```{r readPhy, eval=FALSE}
data("phylo_Furnariidae")
```
An environmental descriptor is also needed to run mcfly function. In this example we use a dataframe present in mcfly package that contains four environmental variables (mean anual temperature, temperature sazonality, mean anual precipitation and precipitation sazonality) and geographic coordinates of sites (latitude and longitude). 
```{r envdata, eval= FALSE}
data("envir")
```

These three datasets comprise the basic information needed to perform the individual-based simulation model presented in mcfly function.

##Data proccessing 

Since large trees require great computational performance and given that the power of analysis is affected by the representativeness of species pool on phylogenetic hypothesis used, we splitted the phylogeny of Furnariidae species, and selected the subfamily Synallaxinae to perform the analysis, since it contains a reasonable number of species widely distributed across sites in the Neotropics.
```{r phylo_process, eval= FALSE}
sub.tree<- ape::subtrees(phylo_Furnariidae) #splitting internal nodes in the phylogeny
n.nodes<-length(sub.tree)
nspp.nodes<-matrix(NA,n.nodes,1)
for(i in 1:n.nodes){
  nspp.nodes[i,]<-sub.tree[[i]]$Ntip #number of species in each internal node
}

dim.comm.subset<-matrix(NA,nspp.nodes,2)
for(j in 1:n.nodes){
  sub.phy<-sub.tree[[j]]
  comm.sub.phy<-Furnariidae[,sub.phy$tip.label]
  zero.row<-which(rowSums(comm.sub.phy)==0)
  comm.subset<-comm.sub.phy[-zero.row,]
  dim.comm.subset[j,]<-dim(comm.subset) # number of sites showing occurrences of species belonging to each internal node
}

```
We chose Synallaxinae subfamily, that presented nine species distributed through 649 communities. Species occurrences were extracted from community data frame. We also filtered environmental variables and geographic coordinates of points to match the occurrences of Synallaxinae species in community data.
```{r comm_procces, eval= FALSE}
plot(sub.tree[[49]]) #Synallaxinae subfamily
sub.phy<- sub.tree[[49]]
comm.sub.phy<- Furnariidae[, sub.tree[[49]]$tip.label] #selecting only sites that present occurrences of species of Synallaxinae
zero.row<- which(rowSums(comm.sub.phy) == 0)
comm.subset<- comm.sub.phy[-zero.row, ] #community matrix with occurrences of Synalaxynae species
coords.subset<- as.matrix(envir[rownames(comm.subset), c(1,2)]) #extracting coordinates
```


##Setting parameters to enter in mcfly function
The parameters that need to be set in mcfly function are defined mainly by the features of species niche positions, which are defined based on a single environmental variable or a factor/principal component summarizing covariation among a number of environmental variables. In this example we decided to use latitude, assuming that this single variable represents a latent variable that summarizes the variation in a set of niche dimensions.
The parameters that must be defined before running mcfly function are:


* **sigma**: 
represents the environmental niche breadth and was set as being the standard deviation of the environmental variable.

* **root.value**:
the trait value at the root of phylogenetic tree, was set as being the mean value of niche dimension of species.

* **theta**:
the optimum value of niche position for the species. We defined this value as being the predicted value derived from a linear regression of diversity (entropy) on the niche variable (latitude) that maximized diversity.

```{r choos_params, eval=FALSE}
env.subset<-envir[rownames(comm.subset),]
envir<- scales::rescale(as.matrix(env.subset$Lat),c(1,100)) #scaling the environmental variable to vary between 1 and 100
sigma<- sd(envir) #standard deviation of environmental variable
root.value<- mean(envir) #mean value
div<- vegan::renyi(comm.subset,scales=1) #diversity measure
mod<- lm(div~envir) #linear model
pred<- predict.lm(mod) #predicted relationship among diversity and the environmental variable
ED<- cbind(envir,pred)
theta<- ED[which.max(pred),1] #optimum niche value
```


Finally, we defined the set of OU alpha parameters that will be used to simulate species traits.

```{r choos_alpha, eval= FALSE}
OU.alpha<- c(0,0.05,0.25,1)
```

We chose the number of runs, that indicates the number of times that the simulation will be repeated, and the number of computer cores to perform the analysis. Here we set the number of cores as being the total number of cores minus 1, but this must be set according to the configuration of your personal computer.
```{r nCores_runs, eval= FALSE}
ncores<- parallel::detectCores() - 1
```

The last parameter to set, **W.r** defines the dispersion limitation of individuals in the simulated landscape used in individual simulation model (to see more details of simulation model used see [MCSim package](http://rstudio-pubs-static.s3.amazonaws.com/159425_80725873417e42fdb13821c10a198281.html). We set W.r value as 0, which generates a scenerio without dispersal limitation. Other dispersal limitation scenarios might be used instead.


Finally, we fed mcfly function with these parameters:
```{r mcfly_run, eval= FALSE}
test.synallaxinae<- mcfly::Mcfly(comm= comm.subset, subset= FALSE, occurrence= TRUE, env= envir, site.coords= coords.subset, tree= sub.phy, OU.alpha= OU.alpha, sigma= sigma, theta= theta, root.value= root.value, runs= 100, ncores= 3, area.m2= 1, m= 0.5, JM= sum(comm.subset), JM.limit= JM,JL= rowSums(comm.subset),nu= 0, speciation.limit= 0, n.timestep= 50,W.r= 0, scenario.ID= "species.sorting", sim.ID= "data", output.dir.path= "OUTPUT_DATA")

```
The output of Mcfly function is a list containing the following objects:

* **Entropy** : 
A data frame containing observed diversity values for entropy 1, 2 and 12 for all communities.

* **Predicted.entropy.1**:
A matrix containing the predicted diversity value for entropy 1 for all the communities accordingly to the four alpha values.

* **$Predicted.entropy.2**: 
A matrix containing the predicted diversity value for entropy 2 for all the communities accordingly to the four alpha values.

* **Predicted.entropy.12**:
A matrix containing the predicted diversity value for entropy 12 for all the communities accordingly to the four alpha values.

* **$K.niche.position**: 
A matrix containing the mean and lower and upper values from 95% confidence interval for K statistic estimated for the four alpha values.

* **$Alpha.niche.position**:
A matrix containing the mean and lower and upper values from 95% confidence interval for alpha statistic estimated for the four alpha values.

* **$Sigsq.niche.position**: 
A matrix containing the mean and lower and upper values from 95% confidence interval for sigma statistic estimated for the four alpha values.

* **Z0.niche.position**: 
A matrix containing the mean and lower and upper values from 95% confidence interval for sigma statistic estimated for the four alpha values.

* **$AIC**: 
A matrix containing the AIC statistic for the three diversity entropy for each alpha value. This table present the following format:


| Alpha         | Ent.1         | Ent.2  | Ent.12 |
|---------------|:-------------:|:------:|:------:|
|alpha=0        | -21.12        | 460    | 913    |
|alpha=0.05     | -165.68       | 279    | 718    |
|alpha=1        | 2580.99       | 1110   | 1280   |

This is the statistic that supports the inference of the most probable strength of the stabilizing selection in the niche  associated with the observed diversity patterns.

* **$W**:
A matrix containing the weights derived from AIC for the three diversity entropy for each alpha value.

* **$R2**: A matrix containing the R^2^ statistic for the three diversity entropy for each alpha value derived form a ordinary linear model.

